---
title: "Exploring bushfire climate drivers with bushfirerisk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring bushfire climate drivers with bushfirerisk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(bushfirerisk)
library(ggplot2)
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

# Introduction

The `bushfirerisk` package provides a cleaned monthly climate dataset (`bushfire_real`) and simple plotting tools to explore environmental drivers of Australian bushfire risk.

The focus is on broad-scale drivers such as temperature, wind, and dryness (low rainfall), rather than the fire behaviour in specific fire events.

This vignette: explains what is in `bushfire_real`, shows how to load and visualise it using `plot_fwi()`, and describes how these variables relate to bushfire danger.

# Dataset overview

The dataset `bushfire_real` is included with the package. It contains monthly averages over southeast Australia (approximately 144°E–155°E, 29°S–40°S) from January 1979 to December 2020.

It has the following variables:

-   `month`: Month of observation (date-time, first day of month).
-   `temp_c`: 2m air temperature in °C (higher temperature → higher fire danger).
-   `wind_ms`: 10m wind speed in m/s (stronger wind → faster fire spread).
-   `dryness_index`: Mean total precipitation from ERA5, in metres. Lower values mean drier conditions.

These are broad climate drivers of fire weather. They are simplified inputs to the Fire Weather Index framework, but this dataset does not calculate a complete FWI score.

# Loading the data

After you load the package, you can access the dataset directly:

```{r}
library(bushfirerisk)

# load the data
data(bushfire_real)

head(bushfire_real)

# basic summary
summary(bushfire_real)
```

# Visualising long-term trends

The package includes a helper plotting function, `plot_fwi()`, which draws a monthly time series for any one of the key variables.

To look at temperature through time:

```{r}
plot_fwi(bushfire_real, var = "temp_c")
```

To look at wind speed:

```{r}
plot_fwi(bushfire_real, var = "wind_ms")
```

To look at dryness:

```{r}
plot_fwi(bushfire_real, var = "dryness_index")
```

# Interpreting plots

-   A long-term upward trend in `temp_c` suggests warmer background

-   conditions that can elevate bushfire risk.

-   Periods of higher `wind_ms` indicate conditions that favour faster,\
    more intense fire spread.

-   Periods where `dryness_index` is very low indicate drier months (less rain), which supports fuel drying.

These ingredients — heat, wind, and dryness — are known to contribute to catastrophic fire weather in southeast Australia.

# Limitations

-   The dataset is spatially averaged over a large region. It is NOT location-specific.

-   Humidity and fuel moisture indices are not included in this version.

-   `plot_fwi()` is a visual aid tool; it is not a predictive fire danger model.

# Summary

In summary:

-   `bushfire_real` gives a 40+ year record (1979–2020) of temperature, wind, and rainfall-derived dryness in southeast Australia.

-   `plot_fwi()` makes it easy to visualise how each driver changes over time.

-   These variables are core components of how bushfire danger is assessed in practice.

This package is for teaching and data exploration in ETC5523. It is simple, clear, and easy to reproduce.
